--- 
author: "Håkon Kaspersen"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
geometry: "left=4cm,right=3cm,top=3cm,bottom=3cm"
subparagraph: true
output:
  bookdown::pdf_book:
    latex_engine: xelatex
    keep_tex: true
    fig_caption: yes
    toc: false
    citation_package: natbib
    includes:
      before_body: frontpage.tex
      after_body: after_body.tex
      in_header: preamble.tex
fontsize: 11pt
linestretch: 1.3
documentclass: book
bibliography: [packages.bib, library.bib]
biblio-style: [plainnatnew.bst]
natbiboptions: "numbers,square,comma,sort,compress"
link-citations: yes
---
\pagestyle{empty}
# Acknowledgements {-}
During my time as a PhD student I've been acquainted with many supportive and fantastic people, both at NVI and the School of Veterinary Medicine at NMBU. I am humbled by all the support and kind words I have recieved during these three years. First and foremost, I would like to thank my main supervisor, Dr. Karin Lagesen, for being available at all times, giving structured feedback, and for sparking my interest for bioinformatics. I would like to thank Dr. Anne Margrete Urdahl for her support and interest, as well as the occasional coffee you brought. Furthermore, I thank Dr. Roger Simm for his knowledge in molecular biology, and the helpful discussions we had about genes and mutations. Dr. Madelaine Norström, your help in extracting data from the internal data system at the institute was invaluable, and your knowledge of epidemiology and statistics have been a pillar in both the isolate selection process and analysis. Lastly, I would like to thank Henning Sørum for his innate curiosity of science and seemingly endless knowledge. 

As for my coworkers at NVI, I would like to show my immense gratitude to Solveig Sølverød Mo, Camilla Sekse and Jannice Schau Slettemeås, who have been amazingly supportive both at a personal and professional level. Thank you for the helpful discussions and coffee breaks, and the general interest you have shown in my work. Your everyday support meant more to me than you may have realized. I would like to extend my gratitute toward the rest of the Research group for Food Safety and Animal Health - thank you for the Friday cake discussions and support. Also, I would like to thank Eve Zeyl Fiskebeck for sharing her knowledge on phylogenetics and evolution to me in an understandable and informative way.

For my colleagues at the School of Veterinary Medicine, I would like to thank Erik Paulshus, Karla Meza, Leif Lotherington, Aud Kari Fauske, Gaute Skogtun, Stanislav Iakhno, Mari Røken, Ane Mohn Bjelland, Cristopher Nilsson and Silje Mogstad Finstad for being amazing coworkers and supporters throughout the years. Thank you for the ski trips, barbecues and cabin trips, and the good work environment.

Lastly, I would like to thank my friends and family for being loving and caring during these years. And a special thank you to my partner for being patient with me and always being understanding and supportive.

`r tufte::quote_footer("--- Håkon Kaspersen, Oslo, 2019")`

# Disclosure {-}
The work of this thesis was carried out at the Norwegian Veterinary Institute (NVI), Oslo, Norway, from 2016 to 2019. The PhD project was part of a project named "Quinolone resistant *E. coli* despite low antimicrobial usage: mechanisms and possible preventive measures", in collaboration with the School of Veterinary Medicine, Norwegian University of Life Sciences, and the private industry.
This thesis was written in RStudio with `r version$version.string` [@R-base] using the Bookdown package [@R-bookdown]. All code is published on github: github.com/hkaspersen/hkaspersen_thesis.

# Funding {-}
This study was funded by the Research Funding for Agriculture and the Food Industry (NFR project 255383).

\tableofcontents

\mainmatter
\pagestyle{fancy}
```{r include=FALSE}
# automatically create a bib database for R packages

packages <- c("bookdown",
              "knitr",
              "rmarkdown",
              "tidyverse",
              "PBSmapping",
              "ggmap",
              "rworldmap",
              "kableExtra",
              "tufte",
              "vegan",
              "adephylo",
              "ggtree",
              "phangorn",
              "Biostrings",
              "kableExtra",
              "kable")

knitr::write_bib(c(.packages(), packages), 'packages.bib')
```

# Summary (English)

The occurrence of antimicrobial resistance in Norwegian livestock is generally low. This is a consequence of good animal health and welfare, which reduces the need for antimicrobial use. Quinolones are important antimicrobial compounds both for human and animal health. An increased occurrence of resistance towards these compounds in livestock have been observed in many European countries. In Norway, the use of quinolones in livestock has been restricted. Prophylactic use of any antimicrobial is prohibited. The Norwegian monitoring programme for antimicrobial resistance in feed, food and animals (NORM-VET) performs surveillance of antimicrobial resistance among commensal _E. coli_. The programme identified a low occurrence of quinolone resistance among several tested animal species. However, quinolone resistant _E. coli_ (QREC) has been detected at low levels in a high proportion of samples from both pigs and broilers. Due to the low quinolone usage, this finding was somewhat surprising. Thus, further investigation of the origin of these bacteria was warranted. The aim of this study was to compile existing data on quinolone resistance occurrence. Furthermore, quinolone resistance mechanism characterization was performed. Relationships between the isolates were investigated by using a high resolution phylogenetic approach. The results showed a low occurrence of QREC among the included animal species. A significantly higher occurrence was observed in broilers. _In silico_ characterization of quinolone resistance mechanisms identified chromosomal mutations as the major resistance determinant. Phylogenetic analysis of QREC provided evidence for dissemination in the broiler and pig production chains. Possible persistence of QREC was detected in the broiler production environment. Major QREC sequence types were detected among the samples from broilers. Some of these sequence types had previously been reported in other Nordic countries. Furthermore, phylogenetic analysis indicate that commensal _E. coli_ rarely develop quinolone resistance in the broiler production environment. These results provide evidence for introduction of QREC to the Norwegian broiler production via imported breeding birds. The results highlight the importance of biosecurity measures at the top of the pyramid, to prevent dissemination of QREC.

# Summary (Norwegian)

Norge har en av de laveste forekomstene av antibiotikaresistens i verden grunnet god dyrehelse og –velferd. Forbruket av kinoloner, som er svært viktige antibiotika for både dyr og mennesker, er svært lavt i Norge, og profylaktisk bruk av antibiotika er forbudt i husdyrproduksjonen.  I mange andre europeiske land er det oppdaget en økende forekomst av kinolonresistente _E. coli_ (QREC), som sannsynligvis er koblet til et økt forbruk av dette antibiotikumet. I Norge derimot viser data fra overvåkningsprogrammet for antibiotikaresistens i mikrober fra fôr, dyr og næringsmidler (NORM-VET) en lav totalforekomst av QREC blant norske husdyr. QREC har likevel blitt detektert fra en stor andel prøver fra gris og slaktekylling, men mengden QREC i hver prøve ser ut til å være lav. Disse funnene førte til spørsmål rundt deres opphav. Målet med dette prosjektet var å sammenfatte eksisterende data på forekomst av QREC i ulike dyrearter i Norge, karakterisere kinolonresistensmekanismer i QREC stammer, samt beskrive forholdet mellom stammene ved bruk av dype fylogenetiske metoder. Resultatene viser en overordnet lav forekomst av QREC, men en signifikant høyere relativ forekomst ble observert i slaktekylling. Kromosomale mutasjoner ble identifisert som hovedmekanisme for den observerte kinolonresistensen. Fylogenetiske analyser av sekvensdataene viste en klonal spredning av QREC i både slaktekyllingproduksjonen og slaktegrisproduksjonen, og mulig persistens av QREC i miljøet der slaktekyllingene oppholder seg. Videre fylogenetisk analyse av både villtype _E. coli_ og QREC viste at villtype _E. coli_ i liten grad utvikler kinolonresistens i slaktekyllingproduksjonen. Resultatene viser at introduksjon av QREC fra importerte foreldredyr er hovedårsaken til den observerte forekomsten av QREC i den norske slaktekyllingproduksjonen. Disse resultatene belyser viktigheten av biosikkerhetstiltak høyere i slaktekyllingpyramiden for å hindre spredning av QREC nedover i produksjonen.

# Abbreviations

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

tibble(
  Term = c("Quinolone resistant E. coli",
           "Antimicrobial resistance",
           "Whole genome sequencing",
           "Open Reading Frame",
           "European Committee on Antimicrobial Susceptibility Testing",
           "Epidemiological cut off",
           "Minimum inhibitory concentration",
           "Maximum likelihood",
           "Quinolone resistance determining region",
           "Antimicrobial susceptibility testing",
           "Plasmid mediated quinolone resistance",
           "Basic Local Alignment Search Tool",
           "High Throughput Sequencing",
           "Single Nucleotide Polymophism",
           "Antimicrobial Resistance Identification By Assembly",
           "Non-Metric Multidimentional Scaling",
           "Sequence Type",
           "Akaike's Information Criteria",
           "Bayesian Information Criteria",
           "Principal Component Analysis"),
  Abbreviation = c("QREC",
                   "AMR",
                   "WGS",
                   "ORF",
                   "EUCAST",
                   "ECOFF",
                   "MIC",
                   "ML",
                   "QRDR",
                   "AST",
                   "PMQR",
                   "BLAST",
                   "HTS",
                   "SNP",
                   "ARIBA",
                   "NMDS",
                   "ST",
                   "AIC",
                   "BIC",
                   "PCA"))%>%
  arrange(Term) %>% # i.e. alphabetical order by Term
  kable(booktab = T) # create nice table
```

# List of publications

**Occurrence of quinolone resistant _E. coli_ originating from different animal species in Norway.**

Håkon Kaspersen, Anne Margrete Urdahl, Roger Simm, Jannice Schau Slettemeås, Karin Lagesen, Madelaine Norström (2018). _Veterinary Microbiology_, 217(February):25-31. (10.1016/j.vetmic.2018.02.022).
\
\
\
**Dissemination of quinolone resistant Escherichia coli in the Norwegian broiler and pig production chain, and possible persistence in the broiler production environment.**

Håkon Kaspersen, Camilla Sekse, Eve Zeyl Fiskebeck, Jannice Schau Slettemeås, Roger Simm, Anne Margrete Urdahl, Madelaine Norström, Karin Lagesen. Submitted to _Applied and Evironmental Microbiology_.
\
\
\
**Comparative genome analyses of wild type- and quinolone resistant _Escherichia coli_ indicate dissemination of QREC in the Norwegian broiler breeding pyramid.**

Håkon Kaspersen, Eve Zeyl Fiskebeck, Camilla Sekse, Jannice Schau Slettemeås, Anne Margrete Urdahl, Madelaine Norström, Karin Lagesen, Roger Simm. Submitted to _Frontiers in Microbiology_.