--- 
author: "Håkon Kaspersen"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
geometry: "left=4cm,right=3cm,top=3cm,bottom=3cm"
subparagraph: true
output:
  bookdown::pdf_book:
    latex_engine: xelatex
    keep_tex: true
    fig_caption: yes
    toc: false
    citation_package: natbib
    includes:
      before_body: frontpage.tex
      after_body: after_body.tex
      in_header: preamble.tex
fontsize: 11pt
linestretch: 1.3
documentclass: book
bibliography: [packages.bib, library.bib]
biblio-style: [plainnatnew.bst]
natbiboptions: "numbers,square,comma,sort,compress"
link-citations: yes
---
\pagestyle{empty}
# Acknowledgements {-}
During my time as a Ph.D. student I've been acquainted with many supportive and fantastic people, both at NVI and the Institute of Veterinary Sciences at NMBU. I am humbled by all the support and kind words I have recieved during these three years. First and foremost, I would like to thank my main supervisor, Dr. Karin Lagesen, for being available at all times, giving structured feedback, and for sparking my interest for bioinformatics. I would like to thank Dr. Anne Margrete Urdahl for her support and interest, as well as the occasional coffee you brought. Furthermore, I thank Dr. Roger Simm for his knowledge in molecular biology, and the helpful discussions we had about genes and mutations. Dr. Madelaine Norström, your help in extracting data from the internal data system at the institute was invaluable, and your knowledge of epidemiology and statistics have been a pillar in both the isolate selection process and analysis. Lastly, I would like to thank Henning Sørum for his innate curiosity of science and seemingly endless knowledge. 

As for my coworkers at NVI, I would like to show my immense gratitute to Solveig Sølverød Mo, Camilla Sekse and Jannice Schau Slettemeås, who have been amazingly supportive both at a personal and professional level. Thank you for the helpful discussions and coffee breaks, and the general interest you have shown in my work. Your everyday support meant more to me than you may have realized. I would like to extend my gratitute toward the rest of the Research group for food safety and animal health - thank you for the friday cake discussions and support. Also, I would like to thank Eve Zeyl Fiskebeck for sharing her knowledge on phylogenetics and evolution to me in an understandable and informative way.

For my colleagues at the Institute of Veterinary Sciences, I would like to thank Erik Paulshus, Karla Meza, Leif Lotherington, Aud Kari Fauske, Gaute Skogtun, Stanislav Iakhno, Mari Røken, Ane Mohn Bjelland, Cristopher Nilsson and Silje Mogstad Finstad for being amazing coworkers and supporters throughout the years. Thank you for the ski trips, barbecues and cabin trips, and the good work environment.

Lastly, I would like to thank my friends and family for being loving and caring during these years. And a special thank you to my partner for being patient with me and always being understanding and supportive.

`r tufte::quote_footer("--- Håkon Kaspersen, Oslo, 2019")`

# Disclosure {-}
The work of this thesis was carried out at the Norwegian Veterinary Institute (NVI), Oslo, Norway, from 2016 to 2019. The Ph. D. project was part of a project named "Quinolone resistant *E. coli* despite low antimicrobial usage: mechanisms and possible preventive measures", in collaboration with the Institute of Veterinary Sciences, Norwegian University of Life Sciences, and the private industry.
This thesis was written in RStudio with `r version$version.string` [@R-base] using the Bookdown package [@R-bookdown]. All code is published on github: www.github.com/hkaspersen.

# Funding {-}
This project was funded by the Norwegian Research Council (NFR no. 255383) and the Norwegian food industry.

\tableofcontents

\mainmatter
\pagestyle{fancy}
```{r include=FALSE}
# automatically create a bib database for R packages

packages <- c("bookdown",
              "knitr",
              "rmarkdown",
              "tidyverse",
              "PBSmapping",
              "ggmap",
              "rworldmap",
              "kableExtra",
              "tufte",
              "vegan",
              "adephylo",
              "ggtree",
              "phangorn",
              "Biostrings",
              "kableExtra",
              "kable")

knitr::write_bib(c(.packages(), packages), 'packages.bib')
```

# Summary (English)

The occurrence of antimicrobial resistance in Norwegian livestock is generally low. This is a consequence of good animal health and welfare, which reduces the need for antimicrobial use. Quinolones are highly important and effective antimicrobial compounds both for human and animal health, and increased occurrence of resistance toward these compounds in livestock have been observed in many European countries. In Norway, the use of quinolones in livestock has been greatly restricted, and prophylactic use of antimicrobials is prohibited. The Norwegian monitoring programme for antimicrobial resistance in feed, food and animals (NORM-VET) do surveillance of antimicrobial resistance among commensal _E. coli_, and identified a low overall occurrence among several tested animal species. However, quinolone resistant _E. coli_ (QREC) has been detected at low levels in a high proportion of samples from both pigs and broilers. This finding was surprising, and further information about these isolates and their origin was of interest. The goal of this study was to compile and compare existing data on quinolone resistance occurrence, characterize quinolone resistance mechanisms in QREC isolates, and use high-resolution phylogenetic methods with whole genome sequencing data to describe the relationship between the isolates to possibly identify their origins. The results showed an overall low occurrence of QREC in animals in Norway, but a significantly higher occurrence was observed in wild birds and broilers. _In-silico_ characterization of quinolone resistance mechanisms identified chromosomal mutations as the predominant resistance determinant. Investigations into the phylogenetic relationships between the isolates provided evidence for a clonal dissemination of QREC isolates in the broiler and pig production chains, and possible persistence in the broiler production environment. However, phylogenetic analysis revealed a temporal re-introduction of major sequence type groups of _E. coli_. The results give strong evidence for (re)introduction of QREC to the Norwegian broiler production via imported breeding birds, and highlight the importance of biosecurity measures at the top of the pyramid to prevent dissemination of QREC isolates down the pyramid.

# Summary (Norwegian)

Et lavt forbruk av antibiotika grunnet god dyrehelse og -velferd har ført til at Norge har en av de laveste forekomstene av antibiotikaresistens i verden. Kinoloner er svært viktige antibiotika både for mennesker og dyr, men en økt forekomst av resistens mot dette antibiotikumet er oppdaget i mange Europeiske land. I Norge er forbruket av dette antibiotikumet svært lavt, og profylaktisk bruk av antibiotika er forbudt. Data fra overvåkningsprogrammet for antibiotikaresistens i mikrober fra fôr, dyr og næringsmidler (NORM-VET) viser en lav totalforekomst av kinolonresistente _E. coli_ (QREC) blant Norske husdyr. QREC har likevel blitt detektert i små mengder i en stor andel av prøver fra gris og slaktekylling. Dette var overraskende funn som førte til spørsmål om deres opphav, og nødvendigheten av mer informasjon om disse bakteriene. Målet med dette prosjektet var å samle og sammenlikne eksisterende data på forekomst av QREC i ulike dyrearter i Norge, karakterisere kinolonresistensmekanismer i QREC stammer, samt beskrive forholdet mellom stammene med dype fylogenetiske metoder for å kunne identifisere deres opphav. Resultatene viser en overordnet lav forekomst av QREC blant Norske dyrearter, men en signifikant høyere relativ forekomst ble observert i villfugl og slaktekylling. Kromosomale mutasjoner ble identifisert som hovedmekanisme for den observerte kinolonresistensen. Fylogenetiske analyser av sekvensdataene ga bevis for en klonal spredning av QREC i både slaktekyllingpyramiden og slaktegrisproduksjonen, og mulig persistens av QREC i slaktekyllingmiljøet. Videre fylogenetisk analyse av både villtype _E. coli_ og QREC viste at både QREC og villtype _E. coli_ blir re-introdusert i slaktekyllingpyramiden. Resultatene gir bevis for at re-introduksjon av QREC fra importerte foreldredyr er hovedårsaken til den observerte forekomsten av QREC i den Norske slaktekyllingproduksjonen. Disse resultatene belyser viktigheten av biosikkerhetstiltak høyere i slaktekyllingpyramiden for å forhindre spredning av QREC nedover i produksjonen.

# Abbreviations

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

tibble(
  Term = c("Quinolone resistant E. coli",
           "Antimicrobial resistance",
           "Whole genome sequencing",
           "Multi drug resistance",
           "Epidemiological cut off",
           "Minimum inhibitory concentration",
           "Locally colinear blocks",
           "Maximum likelihood",
           "Quinolone resistance determining region",
           "Antimicrobial susceptibility testing",
           "Plasmid mediated quinolone resistance"),
  Abbreviation = c("QREC",
                   "AMR",
                   "WGS",
                   "MDR",
                   "ECOFF",
                   "MIC",
                   "LCB",
                   "ML",
                   "QRDR",
                   "AST",
                   "PMQR"))%>%
  arrange(Term) %>% # i.e. alphabetical order by Term
  kable(booktab = T) # create nice table
```

# List of publications

**Occurrence of quinolone resistant _E. coli_ originating from different animal species in Norway.**

Håkon Kaspersen, Anne Margrete Urdahl, Roger Simm, Jannice Schau Slettemeås, Karin Lagesen, Madelaine Norström (2018). _Veterinary Microbiology_, 217(February):25-31. (10.1016/j.vetmic.2018.02.022).
\
\
\
**Phylogenetic analysis indicate dissemination and persistence of quinolone resistant _Escherichia coli_ in the broiler and pig production chain in Norway.**

Håkon Kaspersen, Camilla Sekse, Eve Zeyl Fiskebeck, Jannice Schau Slettemeås, Roger Simm, Anne Margrete Urdahl, Madelaine Norström, Karin Lagesen.
\
\
\
**Comparative genome analyses of wild type- and quinolone resistant _Escherichia coli_ indicate dissemination of QREC in the broiler production pyramid and potential sporadic local resistance development.**

Håkon Kaspersen, Eve Zeyl Fiskebeck, Camilla Sekse, Jannice Schau Slettemeås, Anne Margrete Urdahl, Madelaine Norström, Karin Lagesen, Roger Simm.