--- 
author: "Håkon Kaspersen"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
geometry: "left=4cm,right=3cm,top=3cm,bottom=3cm"
subparagraph: true
output:
  bookdown::pdf_book:
    citation_package: natbib
    includes:
      before_body: frontpage.tex
      in_header: preamble.tex
fontsize: 11pt
linestretch: 1.2
toc-depth: 1
documentclass: book
bibliography: [packages.bib, library.bib]
biblio-style: apalike
link-citations: yes
---

```{r include=FALSE}
# automatically create a bib database for R packages

packages <- c("bookdown","knitr","rmarkdown","tidyverse","kableExtra")

knitr::write_bib(c(.packages(), packages), 'packages.bib')
```

# Acknowledgements
I would like to thank...

# Disclosure
This thesis was written in RStudio with `r version$version.string` using the Bookdown package [@R-bookdown].

# Funding
This project was funded by the Norwegian Research Council (NFR no. 255383) and the Norwegian food industry.

# Summary (English)


# Summary (Norwegian)

# Abbreviations

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

tibble(
  Term = c("Quinolone resistant E. coli",
           "Antimicrobial resistance",
           "Whole genome sequencing",
           "Multi drug resistance",
           "Epidemiological cut off",
           "Quinolone sensitive E. coli"),
  Abbreviation = c("QREC",
                   "AMR",
                   "WGS",
                   "MDR",
                   "ECOFF",
                   "QSEC"))%>%
  arrange(Term) %>% # i.e. alphabetical order by Term
  kable(booktab = T) # create nice table
```

# List of publications

1. **Occurrence of quinolone resistant _E. coli_ originating from different animal species in Norway**. Håkon Kaspersen, Anne Margrete Urdahl, Roger Simm, Jannice Schau Slettemeås, Karin Lagesen, Madelaine Norström (2018). _Veterinary Microbiology_, 217(February):25-31. (10.1016/j.vetmic.2018.02.022).

2. **Closely related quinolone resistant Escherichia coli in different animal species in Norway – part of the normal flora or dissemination events?**
Håkon Kaspersen, Camilla Sekse, Jannice Schau Slettemeås, Roger Simm, Anne Margrete Urdahl, Madelaine Norström, Karin Lagesen.

3. **Broiler study**