--- 
author: "Håkon Kaspersen"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
geometry: "left=4cm,right=3cm,top=3cm,bottom=3cm"
subparagraph: true
output:
  bookdown::pdf_book:
    latex_engine: xelatex
    fig_caption: yes
    toc: false
    citation_package: natbib
    includes:
      before_body: frontpage.tex
      after_body: after_body.tex
      in_header: preamble.tex
fontsize: 11pt
linestretch: 1.2
documentclass: book
bibliography: [packages.bib, library.bib]
biblio-style: [plainnatnew.bst]
natbiboptions: "numbers,square,comma,sort,compress"
link-citations: yes
---
\pagestyle{empty}
# Acknowledgements {-}
During my time as a Ph.D. student I've been acquainted with many supportive and fantastic people, both at NVI and the Institute of Veterinary Sciences at NMBU. I am humbled by all the support and kind words I have recieved during these three years. First and foremost, I would like to thank my main supervisor, Dr. Karin Lagesen, for being available at all times, giving structured feedback, and for sparking my interest for bioinformatics. I would like to thank Dr. Anne Margrete Urdahl for her support and interest, as well as the occasional coffee you brought in times of dire need. Furthermore, I thank Dr. Roger Simm for his knowledge in molecular biology, and the helpful discussions we had about genes and mutations. Dr. Madelaine Norström, your help in extracting data from the internal data system at the institute was invaluable, and your knowledge of epidemiology and statistics have been a pillar in both the isolate selection process and analysis. Lastly, I would like to thank Henning Sørum for his innate curiosity of science and seemingly endless knowledge. 

As for my coworkers at NVI, I would like to show my immense gratitute to Solveig Sølverød Mo, Camilla Sekse and Jannice Schau Slettemeås, who have been amazingly supportive both at a personal and professional level. Thank you for the helpful discussions and coffee breaks, and the general interest you have shown in my work. Your everyday support meant more to me than you may have realized. I would like to extend my gratitute toward the rest of the Research group for food safety and animal health - thank you for the friday cake discussions and support. Also, I would like to thank Eve Zeyl Fiskebeck for disseminating her knowledge on phylogenetics and evolution to me in an understandable and informative way.

For my colleagues at the Institute of Veterinary Sciences, I would like to thank Erik Paulshus, Karla Meza, Leif Lotherington, Aud Kari Fauske, Gaute Skogtun, Stanislav Iakhno, Mari Røken, Ane Mohn Bjelland, Cristopher Nilsson and Silje Mogstad Finstad for being amazing coworkers and supporters throughout the years. Thank you for the ski trips, barbecues and cabin trips, and the good work environment.

Lastly, I would like to thank my friends and family for being loving and caring during these years. And a special thank you to my partner for being patient with me and always being understanding and supportive.

`r tufte::quote_footer("--- Håkon Kaspersen, Oslo, 2019")`

# Disclosure {-}
The work of this thesis was carried out at the Norwegian Veterinary Institute (NVI), Oslo, Norway, from 2016 to 2019. The Ph. D. project was part of a project named "Quinolone resistant *E. coli* despite low antimicrobial usage: mechanisms and possible preventive measures", in collaboration with the Institute of Veterinary Sciences, Norwegian University of Life Sciences, and the private industry.
This thesis was written in RStudio with `r version$version.string` [@R-base] using the Bookdown package [@R-bookdown]. All code is published on github: www.github.com/hkaspersen.

# Funding {-}
This project was funded by the Norwegian Research Council (NFR no. 255383) and the Norwegian food industry.

\tableofcontents

\mainmatter
\pagestyle{fancy}
```{r include=FALSE}
# automatically create a bib database for R packages

packages <- c("bookdown",
              "knitr",
              "rmarkdown",
              "tidyverse",
              "PBSmapping",
              "ggmap",
              "rworldmap",
              "kableExtra",
              "tufte",
              "vegan",
              "adephylo",
              "ggtree",
              "Biostrings",
              "kableExtra",
              "kable")

knitr::write_bib(c(.packages(), packages), 'packages.bib')
```

# Summary (English)
The monitoring programme for antimicrobial resistance in feed, food and animals (NORM-VET) have monitored quinolone resistant _E. coli_ (QREC) in multiple animal species since 2006. Sample material is cultured according to EU regulations, which includes culturing of indicator _E. coli_. The programme has since 2009 indicated a low occurrence of QREC in the tested animal species. However, an increased occurrence in broilers was observed in 2009, which instigated further investigations with selective culturing methods. A selective method was introduced in 2014, and the results revealed a very high occurrence of QREC in broilers. The use of quinolones is known to induce quinolone resistance in commensal _E. coli_. However, the use of quinolones is nearly negligible in Norwegian food producing animals, which raised the question of the origin of these bacteria. The goal of this study was to identify the occurrence of QREC in multiple animal species, characterize genomic factors that may explain the observed resistance, identify possible associations between isolates from various origins, and investigate possible underlying risk factors for the occurrence of QREC. Data from NORM-VET 2006 to 2016 was compiled and analysed to get an overview of the occurrence of QREC in multiple animal species. QREC isolates were whole genome sequenced, and the quinolone resistance mechanisms were characterized. Furthermore, phylogenetic analysis was used to describe the relationship between the QREC isolates. The phylogenetic relationship between QREC and sensitive isolates was also determined, together with the presence of mutations in major mutator genes. The results indicate that there is an interspecies variation in QREC occurrence, which may indicate that other factors than quinolone usage may play a role in the occurrence of QREC. Furthermore, quinolone resistance mechanisms identified in the QREC isolates followed the population structure, and phylogenetically related QREC isolates were identified in samples from broilers and red foxes, and also in samples from pigs. Lastly, the phylogenetic relationship between sensitive _E. coli_ isolates and QREC seem to be separate. Taken together, the results indicate a clonal expansion of closely related QREC isolates through the norwegian broiler and pig production chain. This expansion is worrisome, and measures should be taken to ensure their exclusion from the food production. 

# Summary (Norwegian)

# Abbreviations

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

tibble(
  Term = c("Quinolone resistant E. coli",
           "Antimicrobial resistance",
           "Whole genome sequencing",
           "Multi drug resistance",
           "Epidemiological cut off",
           "Minimum inhibitory concentration",
           "Norwegian Sequencing Centre",
           "Locally colinear blocks",
           "Maximum likelihood"),
  Abbreviation = c("QREC",
                   "AMR",
                   "WGS",
                   "MDR",
                   "ECOFF",
                   "MIC",
                   "NSC",
                   "LCB",
                   "ML"))%>%
  arrange(Term) %>% # i.e. alphabetical order by Term
  kable(booktab = T) # create nice table
```

# List of publications

1. **Occurrence of quinolone resistant _E. coli_ originating from different animal species in Norway**
Håkon Kaspersen, Anne Margrete Urdahl, Roger Simm, Jannice Schau Slettemeås, Karin Lagesen, Madelaine Norström (2018). _Veterinary Microbiology_, 217(February):25-31. (10.1016/j.vetmic.2018.02.022).

2. **Phylogenetic analysis indicate dissemination and persistence of quinolone resistant Escherichia coli in the broiler and pig production chain in Norway**
Håkon Kaspersen, Camilla Sekse, Eve Zeyl Fiskebeck, Jannice Schau Slettemeås, Roger Simm, Anne Margrete Urdahl, Madelaine Norström, Karin Lagesen.

3. **Comparative genome analyses of wild type- and quinolone resistant Escherichia coli indicate dissemination of QREC in the broiler production pyramid and potential sporadic local resistance development**
Håkon Kaspersen, Eve Zeyl Fiskebeck, Camilla Sekse, Jannice Schau Slettemeås, Anne Margrete Urdahl, Madelaine Norström, Karin Lagesen, Roger Simm.